---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# decontX

<!-- badges: start -->
[![R-CMD-check](https://github.com/campbio/decontX/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/campbio/decontX/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/campbio/decontX/graph/badge.svg?token=z5YnsXuWqh)](https://codecov.io/gh/campbio/decontX)
[![Bioconductor build](https://bioconductor.org/shields/build/devel/bioc/decontX.svg)](https://bioconductor.org/checkResults/devel/bioc-LATEST/decontX/)
<!-- badges: end -->

Methods for decontamination of single cell data. This package implements both DecontX (Yang et al., [2020](https://doi.org/10.1186/s13059-020-1950-6)) for single-cell RNA-seq data and DecontPro (Yin et al., [2024](https://doi.org/10.1093/nar/gkad1032)) for single-cell protein expression data.

## Installation Instructions

You can install the package through [Bioconductor](https://bioconductor.org/packages/decontX) with:

```{r eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("decontX")
```

Alternatively, install the development version through [GitHub](https://github.com/campbio/decontX) using the `devtools` package:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("campbio/decontX")
```

## Vignettes

Vignettes are available on [Bioconductor](https://bioconductor.org/packages/decontX).

To build the vignettes when installing from GitHub, add the flag `build_vignettes = TRUE`:

```{r eval=FALSE}
library(devtools)
install_github("campbio/decontX", build_vignettes = TRUE)
```

Then vignettes can be accessed through:

```{r eval=FALSE}
vignette('decontX', package = 'decontX')
vignette('decontPro', package = 'decontX')
```
